---
title: "Lab 09 and 10"
author: "Graham Irons"
date: "3/3/2022"
output: html_document
---

#### Step 1: Set Global Options and Load Packages
```{r setup, warning = FALSE, error = FALSE, include = TRUE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(USAboundaries)
library(sf)
library(ggplot2)
library(ggpubr)
library(wesanderson)

```

#### Step 2: Import Water Use Data; Clean Data
The data for this lab are available from the [United States Geological Survey](https://water.usgs.gov/watuse/data/) for years 1950-2015.

Table 1: “Sets” of data, for which each of the sets have ALL column tags in common. Data from USGS (1950-2015). For Plot 1: Fresh vs Saline Withdrawals for Industrial Surface Water Use.
Bold indicate major differences between report years. 

| year | fresh | saline | state |
|:-------:|:------:|:-----------:|:-----------:|
| 1950-1955 | inpt_wsw_fr  | inpt_wsw_sa | **area** |
| 1960-1980 | oi_wsw_fr | oi_wsw_sa | **area** |
| 1985-1990 | in_wswfr | in_wswsa | **scode** |
| 1995 | in_wsw_fr | in_wsw_sa | **statecode** |
| 2000-2015 | in_wsw_fr | in_wsw_sa |statefips |

Table 2: “Sets” of data, for which each of the sets have ALL column tags in common. Data from USGS (1950-2015). For Plot 2: Geographical Comparison of Total Surface Water Withdrawals for Industrial Sector by State.
Bold indicate major differences between report years. 

| year | fresh | saline | state | latitude and longitude |
|:-------:|:------:|:-----------:|:-----------:|:-----------:|
| 1950-1955 | inpt_wsw_fr  | inpt_wsw_sa | **area** | lat / lng |
| 1960-1980 | oi_wsw_fr | oi_wsw_sa | **area** | lat / lng |
| 1985-1990 | in_wswfr | in_wswsa | **scode** | lat / lng |
| 1995 | in_wsw_fr | in_wsw_sa | **statecode** | lat / lng |
| 2000-2015 | in_wsw_fr | in_wsw_sa |statefips | lat / lng |

```{r, message = FALSE, warning = FALSE, error = FALSE, echo = TRUE}

# Read data for 1950-2015
d_1950 <- lapply(excel_sheets(here("data/us1950.xlsx")),                        #chooses sheet to apply following function to
                 function(x) read_excel(here("data/us1950.xlsx"), skip = 3, sheet = x)) %>% #
          reduce(left_join, by = "Area") %>%                                    #join by area column
          clean_names() %>%                                                     #clean names
          select(1, 5, 6) %>%                                                   #select important columns
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0     

d_1955 <- lapply(excel_sheets(here("data/us1955.xlsx")), 
                 function(x) read_excel(here("data/us1955.xlsx"), skip = 3, sheet = x)) %>% 
          reduce(left_join, by = "Area")%>%                                     #join by area column
          clean_names()%>%                                                      #clean names
          select(1, 6, 7) %>%                                                   #select important columns
          mutate(across(2:3, as.numeric)) %>%                                   #set columns1 as numerics
          replace(is.na(.), 0)                                                  #change NA to 0                                                           

d_1960 <- lapply(excel_sheets(here("data/us1960.xlsx")), 
                 function(x) read_excel(here("data/us1960.xlsx"), skip = 3, sheet = x)) %>% 
          reduce(left_join, by = "Area")%>%                                     #join by area column 
          clean_names() %>%                                                     #clean names
          select(1, 15, 16) %>%                                                 #select important columns
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0

d_1965 <- lapply(excel_sheets(here("data/us1965.xlsx")), 
                 function(x) read_excel(here("data/us1965.xlsx"), skip = 3, sheet = x)) %>% 
          reduce(left_join, by = "Area")%>%                                     #join by area column 
          clean_names() %>%                                                     #clean names
          select(1, 15, 16) %>%                                                 #select important columns
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0

d_1970 <- lapply(excel_sheets(here("data/us1970.xlsx")), 
                 function(x) read_excel(here("data/us1970.xlsx"), skip = 3, sheet = x)) %>% 
          reduce(left_join, by = "Area")%>%                                     #join by area column 
          clean_names() %>%                                                     #clean names
          select(1, 15, 16) %>%                                                 #select important columns 
          slice (1:(n() - 5)) %>%                                               #remove last 5 rows from sheet
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0
          
d_1975 <- lapply(excel_sheets(here("data/us1975.xlsx")), 
                 function(x) read_excel(here("data/us1975.xlsx"), skip = 3, sheet = x)) %>% 
          reduce(left_join, by = "Area")%>%                                     #join by area column 
          clean_names() %>%                                                     #clean names
          select(1, 15, 16) %>%                                                 #select important columns
          mutate(across(2:3, as.numeric)) %>%                                  #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0

d_1980 <- lapply(excel_sheets(here("data/us1980.xlsx")), 
                 function(x) read_excel(here("data/us1980.xlsx"), skip = 3, sheet = x)) %>% 
          reduce(left_join, by = "Area")%>%                                     #join by area column 
          clean_names() %>%                                                     #clean names
          select(1, 15, 16) %>%                                                 #select important columns
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0

d_1985 <- read_delim(here("data/us1985.txt"), delim = "\t") %>% 
          clean_names() %>%                                                     #clean names
          select(4, 41, 42) %>%                                                 #select important columns 
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics 
          replace(is.na(.), 0)                                                  #change NA to 0

d_1990 <- read_excel(here("data/us1990.xls")) %>%                               #load water data
          clean_names()%>%                                                      #clean names
          select(4, 41, 42) %>%                                                 #select important columns 
          slice (1:(n() - 1)) %>%                                               # remove last row from sheet
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0

d_1995 <- read_excel(here("data/us1995.xls")) %>%                               #load water data
          clean_names() %>%                                                     #clean names
          select(3, 65, 66) %>%                                                 #select important columns
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0

d_2000 <- read_excel(here("data/us2000.xls")) %>%                               #load water data
          clean_names() %>%                                                     #clean names
          select(2, 17, 18) %>%                                                 #select important columns
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0

d_2005 <- read_excel(here("data/us2005.xls")) %>%                               #load water data
          clean_names() %>%                                                     #clean names 
          select(2, 28, 29) %>%                                                 #select important columns 
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0

d_2010 <- read_excel(here("data/us2010.xlsx")) %>%                              #load water data
          clean_names() %>%                                                     #clean names 
          select(2, 31, 32) %>%                                                 #select important columns 
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0

d_2015 <- read_excel(here("data/us2015.xlsx"), skip = 1) %>%                    #load water data
          clean_names() %>%                                                     #clean name
          select(2, 31, 32) %>%                                                 #select important columns 
          mutate(across(2:3, as.numeric)) %>%                                   #set columns as numerics
          replace(is.na(.), 0)                                                  #change NA to 0

centroids <- readRDS(here("data/state_centroids.rds"))                          #load centroids data

```


#### Step 3: Organize Data by Water Type (Fresh/Saline)
```{r, message = FALSE, warning = FALSE, error = FALSE, echo = TRUE}
#Plot 1
wu_1950 <- d_1950 %>%                                                           #create object that calls on other
            mutate("State" = area,                                              #rename columns
                   "Fresh" = inpt_wsw_fr,
                   "Saline" = inpt_wsw_sa,
                   "Year" = 1950) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_1955 <- d_1955 %>%                                                           #create object that calls on other
            mutate("State" = area,                                              #rename columns
                   "Fresh" = inpt_wsw_fr,
                   "Saline" = inpt_wsw_sa,
                   "Year" = 1955) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_1960 <- d_1960 %>%                                                           #create object that calls on other
            mutate("State" = area,                                              #rename columns
                   "Fresh" = oi_wsw_fr,
                   "Saline" = oi_wsw_sa,
                   "Year" = 1960) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_1965 <- d_1965 %>%                                                           #create object that calls on other
            mutate("State" = area,                                              #rename columns
                   "Fresh" = oi_wsw_fr,
                   "Saline" = oi_wsw_sa,
                   "Year" = 1965) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_1970 <- d_1970 %>%                                                           #create object that calls on other
            mutate("State" = area,                                              #rename columns
                   "Fresh" = oi_wsw_fr,
                   "Saline" = oi_wsw_sa,
                   "Year" = 1970) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_1975 <- d_1975 %>%                                                           #create object that calls on other
            mutate("State" = area,                                              #rename columns
                   "Fresh" = oi_wsw_fr,
                   "Saline" = oi_wsw_sa,
                   "Year" = 1975) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_1980 <- d_1980 %>%                                                           #create object that calls on other
            mutate("State" = area,                                              #rename columns
                   "Fresh" = oi_wsw_fr,
                   "Saline" = oi_wsw_sa,
                   "Year" = 1980) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_1985 <- d_1985 %>%                                                           #create object that calls on other
            mutate("State" = scode,                                             #rename columns
                   "Fresh" = in_wswfr,
                   "Saline" = in_wswsa,
                   "Year" = 1985) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            group_by(State, Year) %>% 
            summarize(across(1:2, sum)) %>% 
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_1990 <- d_1990 %>%                                                           #create object that calls on other
            mutate("State" = scode,                                             #rename columns
                   "Fresh" = in_wswfr,
                   "Saline" = in_wswsa,
                   "Year" = 1990) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            group_by(State, Year) %>% 
            summarize(across(1:2, sum)) %>% 
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_1995 <- d_1995 %>%                                                           #create object that calls on other
            mutate("State" = state_code,                                              #rename columns
                   "Fresh" = in_wsw_fr,
                   "Saline" = in_wsw_sa,
                   "Year" = 1995) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            group_by(State, Year) %>% 
            summarize(across(1:2, sum)) %>% 
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_2000 <- d_2000 %>%                                                           #create object that calls on other
            mutate("State" = statefips,                                              #rename columns
                   "Fresh" = in_wsw_fr,
                   "Saline" = in_wsw_sa,
                   "Year" = 2000) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            group_by(State, Year) %>% 
            summarize(across(1:2, sum)) %>% 
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_2005 <- d_2005 %>%                                                           #create object that calls on other
            mutate("State" = statefips,                                              #rename columns
                   "Fresh" = in_wsw_fr,
                   "Saline" = in_wsw_sa,
                   "Year" = 2005) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            group_by(State, Year) %>% 
            summarize(across(1:2, sum)) %>% 
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_2010 <- d_2010 %>%                                                           #create object that calls on other
            mutate("State" = statefips,                                              #rename columns
                   "Fresh" = in_wsw_fr,
                   "Saline" = in_wsw_sa,
                   "Year" = 2010) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            group_by(State, Year) %>% 
            summarize(across(1:2, sum)) %>% 
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

wu_2015 <- d_2015 %>%                                                           #create object that calls on other
            mutate("State" = statefips,                                              #rename columns
                   "Fresh" = in_wsw_fr,
                   "Saline" = in_wsw_sa,
                   "Year" = 2015) %>% 
            select("State", "Year", "Fresh", "Saline") %>%                      #select new columns
            group_by(State, Year) %>% 
            summarize(across(1:2, sum)) %>% 
            pivot_longer(cols = 3:4,                                            #pivot and rename
                         names_to = "Type",
                         values_to = "Withdrawals")

```


#### Step 4: Organize and Combine Data for Plotting
```{r, warning = FALSE, message = FALSE, error = TRUE, echo = TRUE}

wu_all <- rbind(wu_1960, wu_1965, wu_1970, wu_1975, wu_1980,                    #combine all years into one 
                wu_1985, wu_1990, wu_1995, wu_2000, wu_2005, wu_2010, wu_2015) %>% #object/omit 1950-55
          filter(!State %in% c("78", "72", "11")) %>%                           #filter out non-states
          replace(is.na(.), 0)                                                  #replace NA with 0

wu_all_total <- wu_all %>%                                                      #create total water use per year
                  group_by(Year) %>%                                            #group by year
                  summarize(across(3:3, sum))                                   #add up all withdrawals by year

wu_all_type <- wu_all %>%                                                       #create water use by water type
                  group_by(Type, Year) %>%                                      #group by type and year
                  summarize(across(2:2, sum))                                   #add up withdrawals

```

#### Step 5: Setup Plot for Timeseries of Industrial Surface Water Withdrawals by Type (Fresh/Saline)
```{r, warning = FALSE, message = FALSE, error = TRUE, echo = TRUE, fig.width = 10, fig.height = 4}

ggplot() +
    geom_col(aes(x = Year,                                                      #bar graph/x-axis
                 y = Withdrawals,                                               #y-axis
                 fill = reorder(Type, Withdrawals)),                            #order data by type/withdrawals
             data = wu_all_type,                                                #set data frame
             position = position_dodge(3.5),                                    #position bars next to each other
             width = 3.5) +                                                     #set distance between bars
    geom_line(aes(x = Year,                                                     #line plot/x-axis
                  y = Withdrawals),                                             #y-axis
              data = wu_all_total,                                              #set data frame
              col = "grey",                                                     #set color
              size = 1) +                                                       #set size
    geom_point(aes(x = Year,                                                    #point plot/x-axis
                   y = Withdrawals),                                            #y-axis
               data = wu_all_total,                                             #set data frame
               col = "grey",                                                    #set color
               size = 2) +                                                      #set size
    ggtitle("Industrial Surface Water Withdrawals by Type from 1960 to 2015") + #create plot title
    scale_x_continuous(breaks = scales::pretty_breaks(n = 14), expand = c(0, 0)) +#bring columns to x axis
    scale_y_continuous(breaks = scales::pretty_breaks(n = 10),                  #adds more tick marks to 1st y-axis
                       labels = scales::comma,                                  #adds commas to numbers
                       limits = c(0, 40000),                                    #sets limits for y-axis
                       sec.axis = sec_axis(trans = ~.*2,                        #divide y data by 2
                                           breaks = scales::pretty_breaks(n = 10), #adds more tick marks to 2nd y-axis
                                           name = "Total Withdrawals (Mgal/day)",#axis label
                                           labels = scales::comma),             #adds commas to numbers
                       expand = c(0, 0)) +                                      #brings columns to y axis
    labs(caption = "Figure 1: Industrial surface water withdrawals by type (fresh/saline) in the USA 1950-2015. 
    Data from USGS (1950-2015); created by Graham Irons.",                      #caption
         x = "Year",                                                            #axis labels
         y = "Sector Withdrawals (Mgal/day)",                                   #axis label
         fill = "") +                                                           #no fill label
    theme_classic() +                                                           #removes background colors
    theme(legend.position = "top",                                              #set legend position
          legend.text = element_text(color = "black", size = 8),                #set legend text color/size
          axis.text = element_text(color = "black", size = 8),                  #set axis text color/size
          axis.title.y.right = element_text(color = "grey"),                    #set 2nd y-axis title text color/size
          axis.text.y.right = element_text(color = "grey"),                     #set 2nd y-axis text color/size
          plot.caption = element_text(hjust = -0.15, size = 10, face = "bold")) +  #set plot position/size/font type
    scale_fill_manual(values = wes_palette("Zissou1", n = 2))                   #love the Wes Anderson color theme
    
  

  
```


#### Step 6: Set up Data for Geospatial Map
```{r, warning = FALSE, message = FALSE, error = TRUE, echo = TRUE}

conus <- us_states() %>%                                                        #assign conus/call us_states
          filter(!state_abbr %in% c("AK", "HI", "PR")) %>%                      #filter outliers
          st_transform(5070) %>%                                                #add lines of parallel to map
          select(state_name, State = state_abbr)                                #select and rename

centroids_spatial <- centroids %>%                                              #create object
                      st_as_sf(coords = c("lat", "lng"),                        #set coordinates as lat/lng 
                               crs = 4326)                                      #convert to sf

us_states <- us_states() %>%                                                    #create object
              mutate("State" = statefp)                                         #change name for inner_join

total <- wu_1960 %>%                                                            #create object 
            rbind(wu_2015, by = "Year") %>%                                     #join 2015 data by year
            mutate(across(4, as.numeric)) %>%                                   #set withdrawals as numeric
            group_by(State, Year) %>%                                           #group for next function
            summarize(across(2, sum)) %>%                                       #add up all withdrawals
            merge(us_states) %>%                                                #add us_states df
            mutate("State" = stusps) %>%                                        #rename
            select(1:3) %>%                                                     #select important columns
            inner_join(conus, by = "State") %>%                                 #add coordinate info
            st_as_sf(crs = 5070)                                                #convert to sf

```

#### Step 7: Setup Geospatial Map of Industrial Surface Water Withdrawals by Type (Fresh/Saline) over 1960 and 2015
```{r, warning = FALSE, message = FALSE, error = TRUE, echo = TRUE, fig.width = 10, fig.height = 13}

ggplot() +
          geom_sf(data = total, aes(fill = Withdrawals), col = "black",         #create map
                  size = 0.1) +
          geom_sf_label(data = total, aes(label = State), size = 2) +           #create map label
          facet_wrap(~Year, nrow = 2, scales = "fixed") +                       #show to maps by year
          labs(caption = "Figure 3: Total industrial surface water withdrawals. Data from USGS (1960); created by Graham Irons.", #captions
               x = "Latitude",                                                  #x caption
               y = "Longitude",                                                 #y caption
               fill = "Withdrawals (Mgal/day)") +                               #set legend caption
          theme(axis.text = element_text(color = "grey", size = 6),             #set axis text, color, and size
                axis.ticks.x = element_blank(),                                 #no x-axis tick marks
                axis.ticks.y = element_blank(),                                 #no y-axis tick marks
                panel.background = element_blank(),                             #no background
                panel.grid.major = element_line(size = 0.1,                     #set grid lines size (lines of parallel)/
                                                linetype = "solid",             #set grid line linetype
                                                colour = "lightgrey"),          #set grid line color
                plot.caption = element_text(hjust = 0,                          #set caption location
                                            color = "black",                    #set caption color
                                            size = 9,                           #set caption size
                                            face = "bold"),                     #set caption font type
                legend.position = "top",                                        #set legend position
                legend.title = element_text(size = 10),                         #set legend title size
                legend.text = element_text(size = 9)) +                         #set legend text size
          scale_fill_gradient(low = "lightskyblue",                             #set high density color
                              high = "navy") +                                  #set low density color
          guides(fill = guide_colourbar(title.position = "top",                 #add legend colorbar/set location
                                        title.hjust = .5,                       #set legend colorbar position
                                        ticks = TRUE,                           #add ticks to legend colorbar
                                        ticks.colour = "white",                 #set legend colorbar tick color
                                        barwidth = 10,                          #set legend colorbar width
                                        barheight = .5))                        #set legend colorbar height


```

#**Q1: What is the take home message specific each figure that you made?**
Figure 1 shows that surface water withdrawals have gone down year over year. Figure 2 shows that surface water withdrawals have decreased in most states, but increased in a few. 

#**Q2: Why did you select the data that you did for plot 1?**
I know that we are slowly running out of surface water so I wanted to see if the degree at which we are withdrawing surface water has increased or decreased over several decades. 

#**Q3: Why did you select the data that you did for plot 2?**
I wanted to see if there was any correlation between west or east coast states and their total surface water withdrawals. 

#**Q4: What is the most surprising outcome of each figure that you made?**
Figure 1: I am surprised that surface water withdrawals have decreased over time. 
Figure 2: I am surprise that the states with higher surface water withdrawals decreased their withdrawals while states with lower initial withdrawals ended up increasing. 

#**Q5: What new functions did you implement to support visualization of the data?**
facet_wrap() and ggtitle()

